---
title: "Introduction to R"
tutorial:
  id: "com.example.tutorials.my-first-tutorial"
  version: 0.7
output: 
  learnr::tutorial:
    progressive: true
  html_document:
      toc: TRUE
runtime: shiny_prerendered
---

```{r, include=F}
knitr::opts_chunk$set(echo = FALSE)
library(learnr)
library(lubridate)
library(RCurl)
library(reshape2)
library(ggplot2)
```

```{r import, include=FALSE}
# Any object created here will be available in global environment. Any objects created within excercise chunks are only available within the chunk

# x <- getURL("https://raw.githubusercontent.com/weinbergerlab/Brazil_state/master/prelog_Brazil_state_processed_data.csv")
#ds1<-read.csv("https://raw.githubusercontent.com/weinbergerlab/Brazil_state/master/prelog_Brazil_state_processed_data.csv")
#saveRDS(ds1,'./Data/brazil_hosp.rds')

ds1 <- readRDS('./Data/brazil_hosp.rds')
ds1<-ds1[,c('age_group','date','J12_18')]
ds1$date <- as.Date(ds1$date)
ds2<-ds1[substr(ds1$age_group,1,2)=='09',]
#Get rid of aggregated state and national levels
ds2<-ds2[substr(ds2$age_group,4,4)!='A',]
ds2<-ds2[substr(ds2$age_group,6,7)!='AA',]

# Reshape into a wide format--1 column per state
ds3<-dcast(ds2, date~age_group, value.var='J12_18')
ds3$date<-as.Date(ds3$date)

ds.n <- dcast(ds2, date~., value.var='J12_18', fun.aggregate = sum)
names(ds.n) <- c('date','cases')
ds.n$date <- as.Date(ds.n$date)

date.index <- ds3$date
ds3<-ds3[,-1]
dimnames(ds3)[[2]]<-paste0('state', substr(dimnames(ds3)[[2]],6,7))

state31 <- ds3[,'state31'] #Creates vector named 'state31'
length.state31<-length(state31)
time.index <- 1:length.state31 
mat1<-cbind(time.index, state31) #combine into matrix

```


# Introduction to R
In this tutorial, we will learn how to manipulate and plot data in R.

## Let's start by looking at some real data

These data on hospitalization due to pneumonia among children <12 months of age in Brazil. The data are at the state level, and each column corresponds to one of 27 states. 

The data are stored as a *matrix*, with 27 columns and 144 rows, representing monthly data for 12 years.

```{r br1a, exercise=TRUE, exercise.eval=FALSE}
head(ds3) # View first 6 rows of data
```

```{r br1b, exercise=TRUE, exercise.eval=FALSE}
dimnames(ds3) # See names of the rows and columns
```

```{r br1c, exercise=TRUE, exercise.eval=FALSE}
str(ds3) # Tells you the structure of the object--what type of variables are in there, dimensions, etc
```

```{r br1d, exercise=TRUE, exercise.eval=FALSE}
dim(ds3) # Get the dimensions of the Brazil data
```

Try to just plot the  data from a single state (plot). To obtain data for a single state, you can refer to the column by the name (exactly as it appears in dimnames) or by position (i.e., 2). Try to plot state 43 from region 4 instead ('09_4_43') instead

```{r br2, exercise=TRUE, exercise.eval=FALSE}
plot(ds3[,'state52'], type='l')
```

## Working with vectors and matrices

In R you can work with individual numbers (ie calculate value of 1+1) with vectors of numbers (ie a variable), or with matrices. Data can also be stored in 'data frames' which is like a matrix but that can store a mix of numeric variables and character variables.

First, let's extract one of the columns from the Brazil pneumonia data, create a time index variable, and join them together into a new matrix.

```{r vectormaker, exercise=TRUE, exercise.eval=FALSE}
state31 <- ds3[,'state31'] # Creates vector named 'state31'
length.state31 <- length(state31)
length.state31
```

```{r vectormakerb, exercise=TRUE, exercise.eval=FALSE}
time.index <- 1:length.state31 # Create an index 1,2,3... that has same length as state31

state31[1:50] # Print first 50 observations
time.index[1:50] # Print first 50 observations
```

Vectors (variables) can be combined together into a matrix or a dataframe. (Data frames have the advantage that they can include both numeric and character variables). 

Try to change the code so that you print first 20 rows of data instead of first 10. 

```{r matrixmaker1, exercise=TRUE, exercise.eval=FALSE}
time.index <- 1:length.state31 # Create an index 1,2,3... that has same length as state31
mat1<-cbind(time.index, state31) # Combine into matrix
mat1[1:10,] # Print mat1, first 10 rows

```

```{r matrixmaker11, exercise=TRUE, exercise.eval=FALSE}
# Print the variable names
dimnames(mat1)
```

```{r matrixmaker12, exercise=TRUE, exercise.eval=FALSE}
# Get number of rows and columns of mat1
nrow(mat1)
```

```{r matrixmaker13, exercise=TRUE, exercise.eval=FALSE}
ncol(mat1)
```

You can refer to and extract columns or rows by their position or by their name. Try to change this to extract the first column.

```{r matrixmaker1a, exercise=TRUE, exercise.eval=FALSE}
state31 <- ds3[,'state31'] # Creates vector named 'state31'
length.state31 <- length(state31)
time.index <- 1:length.state31 # Create an index 1,2,3... that has same length as state31
mat1<-cbind(time.index, state31) # Combine into matrix
mat1


```

```{r matrixmaker1aa, exercise=TRUE, exercise.eval=FALSE}
# Extract  column 2 from the matrix mat1 (first 10 observations)
mat1[1:10,'state31']
mat1[1:10,2]
```

You can subset the matrix by refering to specific columns by name or position. 

Try to also print state 33, and print first 10 rows instead of 20.

```{r matrixmaker1b, exercise=TRUE, exercise.eval=FALSE}
br.subset <- ds3[,c('state31', 'state32')]
br.subset[1:20,]

# Equivalent to: br.subset <- ds3[,c(17,18)]

```

## Make and manipulate matrices

Create an empty matrix (filled with missing values), then fill it in.

```{r matrixMaker, exercise=TRUE, exercise.eval=FALSE}
# Let's make a matrix with same number of rows as the Brazil data and 3 columns
br.rows <- nrow(ds3)
mat2 <- matrix(NA, nrow=br.rows, ncol=3)
mat2[1:20,] # Print first 20 rows
```

This code replaces the first column of the matrix with a vector of observations. 

Try to replace the second column with the observed data from the Brazil states. Let's also try to use rnorm to generate random numbers for column 3.

```{r matrixMaker2, exercise=TRUE, exercise.eval=FALSE}
br.rows <- nrow(ds3)
mat2 <- matrix(NA, nrow=br.rows, ncol=3)
mat2[,1] <- ds3[,'state31']
mat2[1:20,]
```

## Arithmetic in R

Now let's do some simple arithmetic with these variables, and combine them into a matrix. 

We might want to add 2 variables, combine into a matrix with named columns, and print first 20 rows.

```{r aggregateRows, exercise=TRUE, exercise.eval=FALSE}
combo <- ds3[,'state31'] + ds3[,'state33']
df1 <- cbind('state31'= ds3[,'state31'], 'state33'= ds3[,'state33'], 'combo'=combo)
df1[1:20,]
```

Add a constant (0.5) to the 3rd column. Create a new vector that is equal to State31+state33+0.5.

```{r addconstant, exercise=TRUE, exercise.eval=FALSE}
combo <- ds3[,'state31'] + ds3[,'state33']
df1 <- cbind('state31'= ds3[,'state31'], 'state33'= ds3[,'state33'], 'combo'=combo)
comboAdd <- df1[,'combo'] + 0.5 
df1[1:20,'combo'] # Print first 20 rows of variable 'combo'
comboAdd[1:20]
```

We also might want to aggregate across all columns. We can do this with the rowSums function or with the apply function, which repeats each operation by row. Then print first 20 observations.

```{r aggregateRows3, exercise=TRUE, exercise.eval=FALSE}
national <- rowSums(ds3)
national[1:20]

national2 <- apply(ds3,1,sum)  # 1 refers to rows and 2 refers to columns
national2[1:20]
```

When adding together two vectors or adding a vector to a matrix, the dimensions need to matrch (ie the length of the vector needs to be the same as the number of rows in the matrix).

```{r calculator3, exercise=TRUE, exercise.eval=FALSE}
x1 <- ds3[1:100,'state31'] # First 100 observations from state31
x2 <- ds3[1:100,'state33'] # First 100 observations from state 33
x3 <- ds3[1:75,'state35'] # First 75 observations from state 35

x1+x2

x1+x3 # Should give a warning message
```


## Working with dates in R

Working with date variables can be tricky in R. The lubridate package can make your life easier.

```{r dates1, exercise=TRUE, exercise.eval=FALSE}
# To make an empty time sequence using dates
from <- as.Date("2003-01-01") # YYYY-MM-DD
timefill <- seq.Date(from=from, length.out=nrow(ds3), by="week") 
timefill[1:144] # Print first few years
```

If you have a vector of dates in a 'character' format, you need to tell R it is a date using as.Date, and tell it what the original format was. For instance mm/dd/yy has the format "%m/%d/%y".

```{r dates2, exercise=TRUE, exercise.eval=FALSE}
# If date is in format mm/dd/yy
date1 <- c("01/01/98", "02/01/98", "03/01/98")
date2 <- as.Date(as.character(date1), format = "%m/%d/%y") # Note use a lowercase Y with 2 digit year
date2

```

Here the input date is in the format mm/dd/yyyy. We tell R this by using format "%m/%d/%Y" Note the uppercase Y here, which indicates a 4 digit year.

```{r dates3, exercise=TRUE, exercise.eval=FALSE}
# If date is in format mm/dd/yyyy
date1 <- c("01/01/1998", "02/01/1998","03/01/1998")
date2 <- as.Date(as.character(date1), format = "%m/%d/%Y") # Note use an uppercase Y with 4 digit year
date2

```

This is very flexible and can take in most any format. For instance, here we have dd-mm-yy, we tell R it is in the format "%d-%m-%y"

```{r dates4, exercise=TRUE, exercise.eval=FALSE}
# If date is in format dd-mm-yy
date1 <- c("01-01-98", "01-02-98","01-03-98")
date2 <- as.Date(as.character(date1), format = "%d-%m-%y")  
date2

```


## Making useful plots

Let's plot the aggregate national-level data. 

```{r plot1, exercise=TRUE, exercise.eval=FALSE}
head(ds.n)
```

This is a bit ugly
```{r}
ggplot(ds.n, aes(x=date, y=cases)) + #Define the dataset used and the x and y variable names
 geom_line()
```
### Make the plot look nicer
You can use 'themes' to control the look of the plots. theme_classic is a nice clean one. And it is good practice to extend the yaxis to 0 with ylim

```{r}
ggplot(ds.n, aes(x=date, y=cases)) + #Define the dataset used and the x and y variable names
 geom_line() +
 theme_classic() + #removes gray background
  expand_limits(y=0) #extend yaxis to 0

```

### Plot multiple states 

To do this, we will start with a dataset that is in the 'long format'--1 row for each date/state combination.

```{r}
head(ds2)
```
 The code will look the same as above, but now we add a facet_wrap() function, which tells ggplot how to group the data. We will select just 3 states to look at
```{r plot3, exercise=TRUE, exercise.eval=FALSE}
ds.subset <- ds2[ds2$age_group %in% c('09_1_11', '09_1_12', '09_1_13'),]

ggplot(ds.subset, aes(x=date, y=J12_18)) + #Define the dataset used and the x and y variable names
 geom_line() +
 theme_classic() + #removes gray background
  expand_limits(y=0)+ #extend yaxis to 0
  facet_wrap(~age_group , scales='free') 

```



### Now let's plot all of the time series at once 
We can just tweak the code above. Remove the facet wrap statement, and add group= and col= statement to the first line

```{r plot5, exercise=TRUE, exercise.eval=FALSE}
ds.subset <- ds2[ds2$age_group %in% c('09_1_11', '09_1_12', '09_1_13'),]

ggplot(ds.subset, aes(x=date, y=J12_18, group=age_group, col=age_group)) + #Define the dataset used and the x and y variable names
 geom_line() +
 theme_classic() + #removes gray background
  expand_limits(y=0) #extend yaxis to 0
```

## Writing Functions

We might want to perform certain custom operations. We can do this by writing and then running functions.

As a silly example, let's provide the function with 2 numbers and add them together. 

We will create a function call 'my.fun1'. This function has 2 inputs 'a' and 'b'. We will supply numeric values for a and b, which are added together and stored as 'z'. Then it will return z

```{r makefun, exercise=TRUE, exercise.eval=FALSE}

my.fun1 <- function(a,b){
  z <- a + b
  return(z)
}

```

Now run the function. Change the input values of a and b and see what happens
```{r fun2,exercise=T, exercise.eval=FALSE}
#Define the function
my.fun1 <- function(a,b){
  z <- a + b
  return(z)
}

#Call the function
res1 <- my.fun1(a=1, b=1)
print(res1)
```

Now try to add two vectors together (same function, but input vectors instead of single numbers (scalars))
```{r fun3,exercise=TRUE, exercise.eval=FALSE}
#Define the function
my.fun1 <- function(a,b){
  z <- a + b
  return(z)
}

#Call the function
res1 <- my.fun1(a=1:10, b=2:11)
print(res1)
```





## RStudio

Now let's go to RStudio and get familiar with the interface and learn how to set the working directory, and how to install and load packages.